Solution of hydrodynamic equilibrium in black-oil reservoirs: the classical material balance equation

### 2014-09-24 pituba/article/article_spe.tex
new changes

@@

\begin{document}
\begin{frontmatter}
\title{Solution of hydrodynamic equilibrium in black-oil reservoirs: the classical material balance equation}
%\title{Obtaining the classical material balance equation from black-oil equations}
%\journal{Journal of Petroleum Science and Engineering}
%\author{R.B. Piccinini}\ead{rbpiccinini@gmail.com}
%\address{Petrobras - UO-BA/ATP-N/RES. Salvador, Bahia, Brazil.}

@@

Spatially averaged quantities are defined for porosity ($\phi$), saturation ($S_\alpha$), density ($\rho_\alpha$) and mass concentration ($Y_{\alpha,k}$) of a component $k$ in phase $\alpha$.

@@

\begin{equation}\label{eq: MBE}
F_p - F_i= \underbrace{N E_o + mN E_g + W E_w}_{\text{fluid volume expansion } (E)} +\underbrace{V_{p1}-V_{p2}}_{\text{pore volume reduction } (C)}
\end{equation}
or
\begin{equation}\label{eq: MBE_short}
F_p - F_i= E + C \, ,
\end{equation}
where new terms $C$, $E$, $F_i$ and $F_p$ are defined in Table~\ref{tab: MBE}.

@@

Assuming reservoir temperature to be sufficiently above gas critical temperature for negligible variation of the Z-factor with pressure, we use hydrostatic equilibrium again and obtain the derivative of gas density with depth:
\begin{equation}\label{eq: gradgdz}
\frac{1}{\rho_g}\left(\frac{\partial \rho_g}{\partial z}\right)_{T} = \frac{\rho_g \vec{g}}{p_g} = \left(\frac{\mathcal{M}}{ Z R T}\right) g \, .
\end{equation}

@@

Oil and gas densities are thus approximately constant with depth.
The reference depth for computing phase density (and also the formation volume factor) can be any arbitrary depth inside phase volume. Under the approximation of a constant density, it is straightforward to show that $\bar{\rho}_\alpha = \rho_\alpha\left(\bar{p}_\alpha\right)$ and $\bar{p}_\alpha=p_\alpha\left(z^0_\alpha\right)$, where $z^0_\alpha$ is the depth of phase gravity center and $\alpha=w, o, g$. The pressure calculated in the material balance equation corresponds to the local pressure at $z^{ref}$, without fluid motion, and only equals the average reservoir pressure $\bar{p}_\alpha$ if the reference depth is that of the gravity center, $z^{ref}=z^0$.

For the case of an oil reservoir with a gas cap, the reference depth should be defined at the gas-oil contact. The point here is that at gas-oil contact both gas and oil densities are computed using the same pressure, its local value. If the reference depth, common to all phases, is otherwise defined below gas-oil contact, gas density will increase with depth at a higher rate because of the oil hydrostatic gradient,
\begin{equation}
\frac{1}{\rho_g}\left(\frac{\partial \rho_g}{\partial z}\right)_{T,Z} = \frac{\rho_o}{\rho_g}\left(\frac{\mathcal{M}}{Z R T}\right)g  \, ,
\end{equation}
what is $\rho_o / \rho_g$ times greater than the gas hydrostatic gradient used in Eq.~\eqref{eq: gradgdz}.

>>>>

% This work is licensed under a Creative Commons Attribution-NonCommercial-ShareAlike 4.0 International License.

@@

\begin{document}
\begin{frontmatter}
\title{Solution of hydrodynamic equilibrium in black-oil reservoirs: the classical material balance equation}
%\title{Obtaining the classical material balance equation from black-oil equations}
%\journal{Journal of Petroleum Science and Engineering}
%\author{R.B. Piccinini}\ead{rbpiccinini@gmail.com}

@@

Spatially averaged quantities are defined for porosity ($\phi$), phase saturation ($S_\alpha$), density ($\rho_\alpha$) and mass concentration ($Y_{\alpha,k}$) of a component $k$ in phase $\alpha$.

@@

\begin{equation}\label{eq: MBE}
F_p - F_i= \underbrace{N E_o + mN E_g + W E_w}_{\text{fluid volume expansion } (E)} +\underbrace{V_{p1}-V_{p2}}_{\text{pore volume compaction} (C)}
\end{equation}
or
\begin{equation}\label{eq: MBE_short}
F_p - F_i= E + C \, ,
\end{equation}
where new terms $C$, $E$, $F_i$ and $F_p$ are defined in Table~\ref{tab: MBE}.

@@

Assuming reservoir temperature to be sufficiently above gas critical temperature for a negligible variation of the Z-factor with pressure, we use hydrostatic equilibrium again and obtain the derivative of gas density with depth:
\begin{equation}\label{eq: gradgdz}
\frac{1}{\rho_g}\left(\frac{\partial \rho_g}{\partial z}\right)_{T} = \frac{\rho_g \vec{g}}{p_g} = \left(\frac{\mathcal{M}}{ Z R T}\right) g \, .
\end{equation}

@@

Oil and gas densities are thus approximately constant with depth.
The reference depth for computing phase density (and also the formation volume factor) can be any arbitrary depth inside phase volume. Under the approximation of a constant density, it is straightforward to show that $\bar{\rho}_\alpha = \rho_\alpha\left(\bar{p}_\alpha\right)$ and $\bar{p}_\alpha=p_\alpha\left(z^0_\alpha\right)$, where $z^0_\alpha$ is the depth of phase gravity center for $\alpha=w, o, g$. The pressure calculated in the material balance equation corresponds to the local pressure at $z^{ref}$ for negligible hydrodynamic gradients and only equals the average reservoir pressure $\bar{p}_\alpha$ if the reference depth is that of the gravity center, $z^{ref}=z^0$.

For the case of an oil reservoir with a gas cap, the reference depth should be defined at the gas-oil contact. The point here is that at gas-oil contact both gas and oil densities are computed using the same pressure its local value. If the reference depth, common to all phases, is otherwise defined below gas-oil contact, gas density will increase with depth at a higher rate because of the oil hydrostatic gradient,
\begin{equation}
\frac{1}{\rho_g}\left(\frac{\partial \rho_g}{\partial z}\right)_{T,Z} = \frac{\rho_o}{\rho_g}\left(\frac{\mathcal{M}}{Z R T}\right)g  \, ,
\end{equation}
what is $\rho_o / \rho_g$ times greater than the gas hydrostatic gradient used in Eq.~\eqref{eq: gradgdz}.
